Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation

### 2022-04-18 ms/manuscript.Rmd
Update MS
As per editor comments

^4^Department of Botany, National Museum of Nature and Science, Tsukuba, Japan

@@

**Methods:** We compiled a community data set of `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` grid cells (20 km × 20 km each) including `r ncol(comm_ferns)` taxa based on > `r occ_sum$filtered$n_specimens$value %>% plyr::round_any(1000, f = floor) %>% number(big.mark = ",")` specimen records.
We combined the community data with a phylogeny and functional traits to analyze taxonomic, phylogenetic, and functional diversity and modeled biodiversity metrics in response to environmental factors and reproductive mode.
Hierarchical clustering was used to delimit bioregions.
Conservation status and threats were assessed by comparing the overlap of significantly diverse grid cells with conservation zones and range maps of native Japanese deer.

@@

<!-- may include title words -->
**Keywords:** apomixis, biodiversity, biogeography, CANAPE, conservation, ferns, Japan, phylogenetic diversity.

@@

Presumably, phylogeny has not been taken into account more prominently in biodiversity studies mainly because the necessary data (DNA sequences and georeferenced occurrence records) and analytic tools have only become available relatively recently [@Soltis2016a; @Folk2021].
These datasets and tools now make it possible to analyze other dimensions of biodiversity such as phylogenetic diversity [@Faith1992] and phylogenetic endemism [@Rosauer2009].
Better ways to characterize biological regions ("bioregions"; `r ie`, areas defined by their taxonomic composition or evolutionary history) are also available [@Laffan2016; @White2019; @Daru2020a], rather than relying on ad hoc characterizations.
Furthermore, incorporating these two frameworks---the categorization of areas into phyloregions with analysis of over/under dispersion of biodiversity---can provide powerful insights into the processes structuring biodiversity and suggest conservation priorities.

A comprehensive understanding of the relationships between richness and other metrics requires densely sampled data, but such datasets are rare on the regional (country) scale.
The ferns of Japan are excellent model system because they have been the target of intense botanical interest for several decades and are densely sampled [reviewed by @Ebihara2019b].
The ferns of Japan include `r total_fern_taxa` native, nonhybrid taxa (including species and varieties) and hundreds of hybrids [nothotaxa\; @Ebihara2019b].
The availability of detailed distribution data [distribution maps at the ca. 10 km scale for nearly all species\; @Ebihara2016b; @Ebihara2017], trait data [multiple quantitative and qualitative traits compiled for identification of nearly all species\; @Ebihara2019b], and DNA sequences for > 97% of species [@Ebihara2010; @Ebihara2019b\; all coverage statistics exclude nothotaxa] make the ferns of Japan an ideal system for investigating the relationships between, and drivers of, multiple dimensions of biodiversity.

@@

Here, we leveraged this exceptionally rich data set to analyze the geographic distribution of biodiversity in detail.
We asked the following questions in our study system, the ferns of Japan: 1. How is biodiversity distributed? 2. How is biodiversity structured? 3. What environmental and biological factors influence biodiversity? 4. How well is biodiversity protected?

## MATERIALS AND METHODS

@@

## Study Site

@@

We used a list of native, nonhybrid fern specimens deposited at herbaria in Japan to quantify occurrences [@Ebihara2016b; @Ebihara2017; @Ebihara2019b].
We chose to use this over other sources such as the Global Biodiversity Information Facility (GBIF) because of its high quality, which obviates many cleaning steps otherwise needed when working with large, publicly available datasets.
Furthermore, the overall sampling completeness of this dataset is also high (`r s_figure("sampling-curve")` in Appendix S1; see Supplemental Data with this article).
The original list comprises `r occ_sum$unfiltered$n_specimens$value %>% scales::number(big.mark = ",")` specimens representing `r occ_sum$unfiltered$n_taxa$value` terminal taxa at the rank of species, subspecies, form, or variety, excluding non-native taxa and nothotaxa.
We treated infraspecific taxa as distinct species during analysis and hereafter use "taxon" and "species" interchangeably to refer to entities at this rank, unless otherwise indicated.
All names are standardized to a common taxonomy, the Japan Green List v. 1.01 [@Ebihara2019c].

Occurrence records were georeferenced and thoroughly vetted as follows [@Ebihara2016b; @Ebihara2017; @Ebihara2019b].
For specimens that lacked latitude and longitude data, georeferencing was done by mapping collection site names to a set of standard ca. 10 km × 10 km grid squares defined by the Statistics Bureau of Japan (the "second-degree mesh"; http://www.stat.go.jp/english/data/mesh/05.html), and the centroid of the second-degree mesh cell used as the specimen location.
In the case that the collection site could not be mapped to a single second-degree mesh cell, it was excluded.
Next, occurrence maps were generated for each taxon showing presence or absence in each second-degree mesh cell.
In the case that a given taxon appeared insufficiently sampled (`r ie`, not present on the map where it would typically be expected to occur), AE and members of the Nippon Fernist Club (local botanists familiar with the ferns of Japan) searched for additional specimens, which were then added to the list.
The occurrence maps were iteratively refined until the vast majority of known Japanese ferns had been observed across their expected ranges; the resulting set of occurrence records can be considered accurate to ca. 10 km (the grain size of the second-degree mesh map).

We further cleaned the list before analysis by filtering out any occurrences not within the second-degree mesh and removing duplicate collections (`r occ_sum$filtered$n_specimens$value %>% scales::number(big.mark = ",")` specimens, `r occ_sum$filtered$n_taxa$value` taxa after filtering).
Given the high quality of our occurrence data and the fact that automated occurrence cleaning algorithms [`r eg`, CoordinateCleaner\; @Zizka2019] have the potential to erroneously exclude true occurrence points [`r ie`, false positives\; @Zizka2020], we chose not to apply additional automated cleaning steps to our data as is often done with occurrence records obtained from GBIF [`r eg`, @Rice2019a; @Suissa2021].

A necessary step in any analysis of biodiversity is to set the grain size used to accurately define communities (co-occurring species).
Here, one must consider that smaller grain size is needed to detect environmental effects, while larger grain size is needed to ensure adequate species sampling.
We created sets of 10-km, 20-km, 30-km, and 40-km grid cells covering Japan using a Mollweide equal-area projection (size refers to the length of the side of each square grid cell).
At each grain size, species occurrences were converted to a presence–absence community matrix (a species was considered present if at least one specimen was recorded in that grid cell).
We calculated sampling redundancy [1 - Richness/Number of specimens\; @Garcillan2003] to quantify adequacy of sampling.
Preliminary analysis indicated that 20-km grid cells are optimal for our data set; redundancy values greatly increased from 10 to 20 km, but much less improvement as grain size is increased beyond 20 km (`r s_figure("grain-size")`).
Although the grid cells are defined with equal area, actual land area of each grid cell varies due to coastlines.
Strictly filtering out all grid cells with less than complete land area would result in a large loss of data, as Japan has many coastlines and small islands.
Therefore, we instead filtered the grid cells by sampling completeness, removing any grid cells with redundancy < 0.1; this dataset was used for all subsequent analyses.
The R package `r pack("sf")` was used for all GIS analyses [@Pebesma2018].

@@

We used data compiled by @Ebihara2019b, which classifies each fern taxon as sexual, apomictic, mixed (both sexual and apomictic modes known from the same taxon), or unknown.
We calculated the percentage of apomictic taxa in each grid cell as the sum of apomictic and mixed taxa divided by total number of taxa (`r s_figure("envir")`B).

@@

We downloaded climate data at the 2.5 minute scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")` [@R-raster], and extracted four variables related to our research questions as follows: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15) (`r s_figure("envir")`).
We identified the intersection of the climate data at the 2.5-minute scale with the 20-km grid cells, then calculated the mean of each climatic variable for each grid cell.
We calculated area as rolling mean area (km^2^) across 100-km latitudinal bands.
This approach reflects the idea that the amount of area with a given climate regime enclosing each grid cell may influence species richness of that cell [@Fahrig2013], and that using raw island areas would not be appropriate in this context because there are just four very large islands with many much smaller ones located close by (`r figure("japan-map")`).

@@

# Number of full plastome genes in FTOL dataset
n_plastome_genes <- plastome_sampling %>%
  filter(data_type == "full_plastome") %>%
  pull(gene) %>%
  n_distinct() %>%
  number(big.mark = ",")

# Number of fern taxa with full plastome sequences in FTOL dataset
n_plastome_fern_taxa <- plastome_sampling %>%
  filter(data_type == "full_plastome") %>%
  # Exclude outgroups
  anti_join(plastome_og, by = "taxon") %>%
  pull(taxon) %>%
  n_distinct() %>%
  number(big.mark = ",")

# Number of fern taxa in FTOL dataset
n_ftol_fern_taxa <- plastome_sampling %>%
  # Exclude outgroups
  anti_join(plastome_og, by = "taxon") %>%
  pull(taxon) %>%
  n_distinct() %>%
  number(big.mark = ",")

# Number of taxa in Japan fern rbcL dataset
n_japan_fern_taxa <- japan_rbcL %>%
  rownames %>%
  n_distinct() %>%
  number(big.mark = ",")

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]; these sequences originate from samples that were identified using the same taxonomic system as the occurrence data [@Ebihara2010; @Ebihara2019b], so we used these preferentially over other data available on GenBank that might include misidentifications or different taxon concepts.
However, using this dataset alone for phylogenetic analysis suffers two drawbacks: it is not densely sampled enough to resolve some internal nodes, and it lacks many lineages that have fossils available for molecular dating.
Furthermore, use of community phylogenies generated by only by sampling the species present in the local community has been shown to produce spurious results in simulation studies [@Park2018].
Because one goal of the study is to understand the distribution of paleo- vs. neoendemic areas (defined in units of time), we require an ultrametric tree.
Therefore, to obtain a robust, ultrametric tree, we combined the *rbcL* data of @Ebihara2019b (`r n_japan_fern_taxa` taxa) with a globally sampled plastid dataset including all fern sequences on GenBank for four widely sequenced plastid genes (*atpA*, *atpB*, *rbcL*, and *rps4*; `r n_ftol_fern_taxa` taxa) and `r n_plastome_genes` other coding, single-copy plastid genes extracted from `r n_plastome_fern_taxa` complete fern plastomes [FTOL v0.0.1\; @Nitta2022b] (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` operational taxonomic units [OTUs] total, including outgroup taxa).
This gene sampling is comparable to a recent global fern phylogeny that resolved relationships across ca. 4,000 taxa using six plastid markers [@Testo2016a], and the addition of plastome data can be expected to increase support along the backbone.

We conducted maximum-likelihood phylogenetic analysis with `r software("IQ-TREE")` v1.6.12 [@Nguyen2015] under the GTR+I+G model (all sites concatenated into a single data matrix), and assessed node support with 1,000 ultrafast bootstrap replicates [@Minh2013].
Molecular dating was conducted with the program `r software("treePL")` v1.0 [@Smith2012] using `r n_fossil_points` fossil calibration points after @Testo2016a, with the exception of treating fossil *Kuylisporites* as belonging to crown *Cyathea*, not *Cyathea* + *Alsophila* [@Loiseau2020].
The root of the tree (most recent common ancestor of all land plants) was fixed to 475 mya [@Testo2016a].
We trimmed the resulting ultrametric tree to only Japanese taxa.
`r software("treePL")` requires branch lengths to be >0, and it sets extremely small branches to an arbitrary minimum positive length.
This behavior may artificially add branch length between taxa with identical DNA sequences, which occur multiple times in the Japanese fern *rbcL* dataset.
We converted any clades in the dated tree consisting of identical sequences to polytomies with the node at 0 million years ago (i.e., the present; Ma).
This tree was used for all subsequent analysis unless indicated otherwise.

@@

We measured phylogenetic diversity (PD) using the Faith's PD [the total branch length connecting all terminal taxa in a community, including the root of the phylogeny\; @Faith1992].
To detect areas with unusually long or short branches, we calculated relative phylogenetic diversity (RPD), which is a ratio of observed PD to PD measured on a tree where all branch lengths have been transformed to equal length [@Mishler2014].

@@

We tested whether centers of endemism have an over-representation of old (paleoendemic) or new (neoendemic) lineages using categorical analysis of neo- and paleoendemism (CANAPE), which involves measuring PE with an alternate tree where all branch lengths are equal, then comparing the ratio between raw PE and alternative PE (RPE) [@Mishler2014].
Low RPE indicates short range-restricted branches (neoendemism) and high RPE indicates long range-restricted branches (paleoendemism).
Since all of these measures are affected by species richness, categorization of endemism is done using the rank of each observed variable relative to those calculated for the null distribution. All measures of biodiversity and CANAPE were carried out using the R package `r pack("canaper")` [@Nitta2022].

It is possible that endemism may be affected by species that are actually wide-ranging, but only have a small range in Japan (a "border effect").
This is especially expected for southern, subtropical islands, which harbor many tropical species with small ranges in Japan but wider ranges outside of the country.
To partially account for the border effect, we also calculated all endemism-related metrics using a data set including species found only in Japan.

@@

Two grid cells lacking environmental data, one grid cell occupied by a single species lacking reproductive mode data, and an outlier with 100% apomictic taxa due to extremely low species richness were excluded from modeling analyses (Appendix S2; see Supplemental Data with this article).

We tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and percentage apomictic taxa [hereafter, % apomictic taxa]) using Moran's I with the `r func("moran.mc")` function in the R package `r pack("spdep")` [@Bivand2013].
In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05) (`r s_table("moran")`).
Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the R package `r pack("SpatialPack")` [@SpatialPack2020].
This test indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature (`r s_table("corr")`), so for the main modeling analysis we only included the following environmental variables: mean annual temperature, precipitation, and precipitation seasonality.

@@

## Datasets

@@

### Specimens (filtered occurrence data) ###
n_specimens <- occ_sum$filtered$n_specimens$value %>%
  scales::number(big.mark = ",")
n_specimens_taxa <- occ_sum$filtered$n_taxa$value
n_specimens_taxa_percent <- n_specimens_taxa %>%
  magrittr::divide_by(total_fern_taxa) %>%
  scales::percent(accuracy = 0.01)

@@

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent` of the total number of native, nonhybrid ferns [`r total_fern_taxa` taxa]) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families.
The topology generally agreed with earlier, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a\; `r s_figure("tree")`].
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* [families follow @PteridophytePhylogenyGroupI2016].
The age of the fern clade is estimated to be `r phylogeny_age` Ma.

The cleaned list of specimens included `r n_specimens` specimens representing `r n_specimens_taxa` taxa (`r n_specimens_taxa_percent`).
The distribution of collection locations was highly skewed, with > 3000 specimens per grid cell northwest of Tokyo in Saitama Prefecture, and many fewer elsewhere (`r s_figure("abundance")`A).
Despite this skew, sampling redundancy was generally good [> 0.3\; @Gonzalez-Orozco2014b] throughout the country (mean `r mean_redundancy`, final community data matrix; SDs are given unless mentioned otherwise; `r s_figure("abundance")`B).
The final community data matrix included `r ncol(comm_ferns)` taxa and `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` grid cells. `r n_excluded_cells` grid cells were excluded due to low redundancy (< 0.1); these were mostly from Hokkaido and coastal areas.

After removing correlated and low-variance traits, the final trait set included `r n_binary_traits` binary trait, `r n_cont_traits` continuous traits, and `r n_qual_traits` qualitative traits (coded as `r n_qual_traits_as_binary` binary traits; Tables `r table_num("traits-used")`, `r s_table_num("phy-sig-binary")`).
The trait matrix included `r n_taxa_traits_sampled` taxa (`r percent_taxa_traits_sampled`) and `r fern_traits %>% select(-taxon) %>% ncol()` traits, with `r percent_traits_missing` missing data.

@@

All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and $\lambda$: all $\lambda$ values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.1, indicating less phylogenetic signal than expected under BM (`r s_table("phy-sig")`).
`r n_d_signif` out of `r nrow(binary_sig_results)` of binary traits (including qualitative traits coded as binary) showed evidence of phylogenetic signal (*D* ≤ 0.5, significantly different from random; `r s_table("phy-sig-binary")`).

@@

Taxonomic richness is lowest on the northernmost island of Hokkaido, then increases with decreasing latitude until reaching a maximum in southern Kyushu, then decreases again in smaller islands further south (`r figure("raw-diversity")`A, `r s_figure("raw-lat")`A).
Observed values of PD and FD were highly correlated with richness as expected (`r figure("raw-diversity")`B, C), showing an initial steep slope that gradually starts to level off at higher richness (`r s_figure("raw-biplot")`A, C).
Observed values of RPD and RFD were not linearly related to richness; however, they showed high variance at low species richness, and low variance at high species richness (`r s_figure("raw-biplot")`B, D).

@@

Our study integrates diverse, thoroughly collected datasets to characterize the spatial biodiversity of Japanese ferns and gain insight into the processes generating these patterns.
We conclude by summarizing some caveats that must be kept in mind when interpreting our results, while also suggesting avenues for future research.

>>>>

^3^Department of Botany, National Museum of Nature and Science, Tsukuba, Japan

@@

**Methods:** We compiled a community data set of `r nrow(comm_ferns) %>% number()` grid cells (20 × 20 km each) including `r ncol(comm_ferns)` taxa based on > `r occ_sum$filtered$n_specimens$value %>% plyr::round_any(1000, f = floor) %>% number()` specimen records.
We combined the community data with a phylogeny and functional traits to analyze taxonomic, phylogenetic, and functional diversity and modeled biodiversity metrics in response to environmental factors and reproductive mode.
Hierarchical clustering was used to delimit bioregions.
Conservation status and threats were assessed by comparing the overlap of significantly diverse grid cells with conservation zones and range maps of native Japanese deer.

@@

<!-- may include title words -->
**KEYWORDS**  
apomixis, biodiversity, biogeography, CANAPE, conservation, ferns, Japan, phylogenetic diversity

@@

Presumably, phylogeny has not been taken into account more prominently in biodiversity studies mainly because the necessary data (DNA sequences and georeferenced occurrence records) and analytic tools have only become available relatively recently [@Soltis2016a; @Folk2021].
These data sets and tools now make it possible to analyze other dimensions of biodiversity such as phylogenetic diversity [@Faith1992] and phylogenetic endemism [@Rosauer2009].
Better ways to characterize biological regions ("bioregions"; `r ie`, areas defined by their taxonomic composition or evolutionary history) are also available [@Laffan2016; @White2019; @Daru2020a], rather than relying on ad hoc characterizations.
Furthermore, incorporating these two frameworks---the categorization of areas into phyloregions with analysis of over/under dispersion of biodiversity---can provide powerful insights into the processes structuring biodiversity and suggest conservation priorities.

A comprehensive understanding of the relationships between richness and other metrics requires densely sampled data, but such data sets are rare on the regional (country) scale.
The ferns of Japan are excellent model system because they have been the target of intense botanical interest for several decades and are densely sampled [reviewed by @Ebihara2019b].
The ferns of Japan include `r total_fern_taxa` native, nonhybrid taxa (including species and varieties) and hundreds of hybrids [nothotaxa\; @Ebihara2019b].
The availability of detailed distribution data [distribution maps at the ca. 10 km scale for nearly all species\; @Ebihara2016b; @Ebihara2017], trait data [multiple quantitative and qualitative traits compiled for identification of nearly all species\; @Ebihara2019b], and DNA sequences for > 97% of species [@Ebihara2010; @Ebihara2019b\; all coverage statistics exclude nothotaxa] make the ferns of Japan an ideal system for investigating the relationships between, and drivers of, multiple dimensions of biodiversity.

@@

Here, we leveraged this exceptionally rich data set to analyze the geographic distribution of biodiversity in detail.
We asked the following questions in our study system, the ferns of Japan: (1) How is biodiversity distributed? (2) How is biodiversity structured? (3) What environmental and biological factors influence biodiversity? (4) How well is biodiversity protected?

# MATERIALS AND METHODS

@@

## Study site

@@

We used a list of native, nonhybrid fern specimens deposited at herbaria in Japan to quantify occurrences [@Ebihara2016b; @Ebihara2017; @Ebihara2019b].
We chose to use this over other sources such as the Global Biodiversity Information Facility (GBIF) because of its high quality, which obviates many cleaning steps otherwise needed when working with large, publicly available data sets.
Furthermore, the overall sampling completeness of this data set is also high (`r s_figure("sampling-curve")` in Appendix S1).
The original list comprises `r occ_sum$unfiltered$n_specimens$value %>% number()` specimens representing `r occ_sum$unfiltered$n_taxa$value` terminal taxa at the rank of species, subspecies, form, or variety, excluding non-native taxa and nothotaxa.
We treated infraspecific taxa as distinct species during analysis and hereafter use "taxon" and "species" interchangeably to refer to entities at this rank, unless otherwise indicated.
All names are standardized to a common taxonomy, the Japan Green List v. 1.01 [@Ebihara2019c].

Occurrence records were georeferenced and thoroughly vetted as follows [@Ebihara2016b; @Ebihara2017; @Ebihara2019b].
For specimens that lacked latitude and longitude data, georeferencing was done by mapping collection site names to a set of standard ca. 10 × 10 km grid squares defined by the Statistics Bureau of Japan (the "second-degree mesh"; http://www.stat.go.jp/english/data/mesh/05.html), and the centroid of the second-degree mesh cell used as the specimen location.
In the case that the collection site could not be mapped to a single second-degree mesh cell, it was excluded.
Next, occurrence maps were generated for each taxon showing presence or absence in each second-degree mesh cell.
In the case that a given taxon appeared insufficiently sampled (`r ie`, not present on the map where it would typically be expected to occur), AE and members of the Nippon Fernist Club (local botanists familiar with the ferns of Japan) searched for additional specimens, which were then added to the list.
The occurrence maps were iteratively refined until the vast majority of known Japanese ferns had been observed across their expected ranges; the resulting set of occurrence records can be considered accurate to ca. 10 km (the grain size of the second-degree mesh map).

We further cleaned the list before analysis by filtering out any occurrences not within the second-degree mesh and removing duplicate collections (`r occ_sum$filtered$n_specimens$value %>% number()` specimens, `r occ_sum$filtered$n_taxa$value` taxa after filtering).
Given the high quality of our occurrence data and the fact that automated occurrence cleaning algorithms [`r eg`, CoordinateCleaner\; @Zizka2019] have the potential to erroneously exclude true occurrence points [`r ie`, false positives\; @Zizka2020], we chose not to apply additional automated cleaning steps to our data as is often done with occurrence records obtained from GBIF [`r eg`, @Rice2019a; @Suissa2021].

A necessary step in any analysis of biodiversity is to set the grain size used to accurately define communities (co-occurring species).
Here, one must consider that smaller grain size is needed to detect environmental effects, while larger grain size is needed to ensure adequate species sampling.
We created sets of 10-km, 20-km, 30-km, and 40-km grid cells covering Japan using a Mollweide equal-area projection (size refers to the length of the side of each square grid cell).
At each grain size, species occurrences were converted to a presence–absence community matrix (a species was considered present if at least one specimen was recorded in that grid cell).
We calculated sampling redundancy [1 – Richness/Number of specimens\; @Garcillan2003] to quantify adequacy of sampling.
Preliminary analysis indicated that 20-km grid cells are optimal for our data set; redundancy values greatly increased from 10 to 20 km, but much less as grain size is increased beyond 20 km (Appendix S1: `r s_figure("grain-size")`).
Although the grid cells are defined with equal area, actual land area of each grid cell varies due to coastlines.
Strictly filtering out all grid cells with less than complete land area would result in a large loss of data, as Japan has many coastlines and small islands.
Therefore, we instead filtered the grid cells by sampling completeness, removing any grid cells with redundancy < 0.1; this data set was used for all subsequent analyses.
The R package `r pack("sf")` was used for all GIS analyses [@Pebesma2018].

@@

We used data compiled by @Ebihara2019b, which classifies each fern taxon as sexual, apomictic, mixed (both sexual and apomictic modes known from the same taxon), or unknown.
We calculated the percentage of apomictic taxa in each grid cell as the sum of apomictic and mixed taxa divided by total number of taxa (Appendix S1: `r s_figure("envir")`B).

@@

We downloaded climate data at the 2.5 minute-scale from the WorldClim database using the `r func("getData")` function in the R package `r pack("raster")` [@R-raster] and extracted four variables related to our research questions: mean annual temperature (bio1; hereafter "temperature"), temperature seasonality (bio4), annual precipitation (bio12; hereafter "precipitation"), and precipitation seasonality (bio15) (Appendix S1: `r s_figure("envir")`).
We identified the intersection of the climate data at the 2.5-minute scale with the 20-km grid cells, then calculated the mean of each climatic variable for each grid cell.
We calculated area as rolling mean area (km^2^) across 100-km latitudinal bands.
This approach reflects the idea that the amount of area with a given climate regime enclosing each grid cell may influence species richness of that cell [@Fahrig2013] and that using raw island areas would not be appropriate in this context because there are just four very large islands with many much smaller ones close by (`r figure("japan-map")`).

@@

# Number of full plastome genes in FTOL dataset
n_plastome_genes <- plastome_sampling %>%
  filter(data_type == "full_plastome") %>%
  pull(gene) %>%
  n_distinct() %>%
  number()

# Number of fern taxa with full plastome sequences in FTOL dataset
n_plastome_fern_taxa <- plastome_sampling %>%
  filter(data_type == "full_plastome") %>%
  # Exclude outgroups
  anti_join(plastome_og, by = "taxon") %>%
  pull(taxon) %>%
  n_distinct() %>%
  number()

# Number of fern taxa in FTOL dataset
n_ftol_fern_taxa <- plastome_sampling %>%
  # Exclude outgroups
  anti_join(plastome_og, by = "taxon") %>%
  pull(taxon) %>%
  n_distinct() %>%
  number()

# Number of taxa in Japan fern rbcL dataset
n_japan_fern_taxa <- japan_rbcL %>%
  rownames %>%
  n_distinct() %>%
  number()

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]; these sequences originate from samples that were identified using the same taxonomic system as the occurrence data [@Ebihara2010; @Ebihara2019b], so we used these preferentially over other data available on GenBank that might include misidentifications or different taxon concepts.
However, using this data set alone for phylogenetic analysis suffers two drawbacks: it is not densely sampled enough to resolve some internal nodes, and it lacks many lineages that have fossils available for molecular dating.
Furthermore, use of community phylogenies generated by only by sampling the species present in the local community has been shown to produce spurious results in simulation studies [@Park2018].
Because one goal of the study is to understand the distribution of paleo- vs. neoendemic areas (defined in units of time), we require an ultrametric tree.
Therefore, to obtain a robust, ultrametric tree, we combined the *rbcL* data of @Ebihara2019b (`r n_japan_fern_taxa` taxa) with a globally sampled plastid data set including all fern sequences on GenBank for four widely sequenced plastid genes (*atpA*, *atpB*, *rbcL*, and *rps4*; `r n_ftol_fern_taxa` taxa) and `r n_plastome_genes` other coding, single-copy plastid genes extracted from `r n_plastome_fern_taxa` complete fern plastomes [FTOL v0.0.1\; @Nitta2022b] (`r ape::Ntip(plastome_tree) %>% number()` operational taxonomic units [OTUs] total, including outgroup taxa).
This gene sampling is comparable to a recent global fern phylogeny that resolved relationships across `r ca` `r number(4000)` taxa using six plastid markers [@Testo2016a], and the addition of plastome data can be expected to increase support along the backbone.

We conducted maximum-likelihood phylogenetic analysis with `r software("IQ-TREE")` v1.6.12 [@Nguyen2015] under the GTR+I+G model (all sites concatenated into a single data matrix), and assessed node support with `r number(1000)` ultrafast bootstrap replicates [@Minh2013].
Molecular dating was conducted with the program `r software("treePL")` v1.0 [@Smith2012] using `r n_fossil_points` fossil calibration points after @Testo2016a, with the exception of treating fossil *Kuylisporites* as belonging to crown *Cyathea*, not *Cyathea* + *Alsophila* [@Loiseau2020].
The root of the tree (most recent common ancestor of all land plants) was fixed at 475 million years ago [Ma\; @Testo2016a].
We trimmed the resulting ultrametric tree to only Japanese taxa.
`r software("treePL")` requires branch lengths to be > 0, and it sets extremely small branches to an arbitrary minimum positive length.
This behavior may artificially add branch length between taxa with identical DNA sequences, which occur multiple times in the Japanese fern *rbcL* data set.
We converted any clades in the dated tree consisting of identical sequences to polytomies with the node at 0 Ma (i.e., the present).
This tree was used for all subsequent analysis unless indicated otherwise.

@@

We measured phylogenetic diversity (PD) using Faith's PD [the total branch length connecting all terminal taxa in a community, including the root of the phylogeny\; @Faith1992].
To detect areas with unusually long or short branches, we calculated relative phylogenetic diversity (RPD), which is a ratio of observed PD to PD measured on a tree where all branch lengths have been transformed to equal length [@Mishler2014].

@@

We tested whether centers of endemism have an over-representation of old (paleoendemic) or new (neoendemic) lineages using categorical analysis of neo- and paleoendemism (CANAPE), which involves measuring PE with an alternate tree where all branch lengths are equal, then comparing the ratio between raw PE and alternative PE (RPE) [@Mishler2014].
Low RPE indicates short range-restricted branches (neoendemism), and high RPE indicates long range-restricted branches (paleoendemism).
Since all of these measures are affected by species richness, categorization of endemism is done using the rank of each observed variable relative to those calculated for the null distribution. All measures of biodiversity and CANAPE were carried out using the R package `r pack("canaper")` [@Nitta2022].

It is possible that endemism may be affected by species that are actually wide-ranging, but only have a small range in Japan (a "border effect").
The border effect is especially expected for southern, subtropical islands, which harbor many tropical species with small ranges in Japan but wider ranges outside of the country.
To partially account for the border effect, we also calculated all endemism-related metrics using a data set including species found only in Japan.

@@

Two grid cells lacking environmental data, one grid cell occupied by a single species lacking reproductive mode data, and an outlier with 100% apomictic taxa due to extremely low species richness were excluded from modeling analyses (Appendix S2).

We tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and percentage apomictic taxa [hereafter, % apomictic taxa]) using Moran's I with the `r func("moran.mc")` function in the R package `r pack("spdep")` [@Bivand2013].
In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05) (Appendix S1: `r s_table("moran")`).
Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the R package `r pack("SpatialPack")` [@SpatialPack2020].
This test indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature (Appendix S1: `r s_table("corr")`), so for the main modeling analysis, we only included the following environmental variables: mean annual temperature, precipitation, and precipitation seasonality.

@@

## Data sets

@@

### Specimens (filtered occurrence data) ###
n_specimens <- occ_sum$filtered$n_specimens$value %>%
  number()
n_specimens_taxa <- occ_sum$filtered$n_taxa$value
n_specimens_taxa_percent <- n_specimens_taxa %>%
  magrittr::divide_by(total_fern_taxa) %>%
  scales::percent(accuracy = 0.01)

@@

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent` of the total number of native, nonhybrid ferns [`r total_fern_taxa` taxa]) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families.
The topology generally agreed with earlier, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a\; Appendix S1: `r s_figure("tree")`].
Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* [families follow @PteridophytePhylogenyGroupI2016].
The age of the fern clade is estimated to be `r phylogeny_age` Ma.

The cleaned list of specimens included `r n_specimens` specimens representing `r n_specimens_taxa` taxa (`r n_specimens_taxa_percent`).
The distribution of collection locations was highly skewed, with > `r number(3000)` specimens per grid cell northwest of Tokyo in Saitama Prefecture, and many fewer elsewhere (Appendix S1: `r s_figure("abundance")`A).
Despite this skew, sampling redundancy was generally good [> 0.3\; @Gonzalez-Orozco2014b] throughout the country (mean `r mean_redundancy`, final community data matrix; SDs are given unless mentioned otherwise; Appendix S1: `r s_figure("abundance")`B).
The final community data matrix included `r ncol(comm_ferns)` taxa and `r nrow(comm_ferns) %>% number()` grid cells after `r n_excluded_cells` grid cells (mostly from Hokkaido and coastal areas) were excluded due to low redundancy (< 0.1).

After removing correlated and low-variance traits, the final trait set included `r n_binary_traits` binary trait, `r n_cont_traits` continuous traits, and `r n_qual_traits` qualitative traits (coded as `r n_qual_traits_as_binary` binary traits; Table `r table_num("traits-used")`; Appendix S1: `r s_table("phy-sig-binary")`).
The trait matrix included `r n_taxa_traits_sampled` taxa (`r percent_taxa_traits_sampled`) and `r fern_traits %>% select(-taxon) %>% ncol()` traits, with `r percent_traits_missing` missing data.

@@

All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and $\lambda$: all $\lambda$ values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.1, indicating less phylogenetic signal than expected under BM (Appendix S1: `r s_table("phy-sig")`).
`r n_d_signif` of `r nrow(binary_sig_results)` of binary traits (including qualitative traits coded as binary) showed evidence of phylogenetic signal (*D* ≤ 0.5, significantly different from random; Appendix S1: `r s_table("phy-sig-binary")`).

@@

Taxonomic richness is lowest on the northernmost island of Hokkaido, then increases with decreasing latitude until reaching a maximum in southern Kyushu, then decreases again in smaller islands further south (`r figure("raw-diversity")`A; Appendix S1: `r s_figure("raw-lat")`A).
Observed values of PD and FD were highly correlated with richness as expected (`r figure("raw-diversity")`B, C), showing an initial steep slope that gradually starts to level off at higher richness (Appendix S1: `r s_figure("raw-biplot")`A, C).
Observed values of RPD and RFD were not linearly related to richness; however, they showed high variance at low species richness, and low variance at high species richness (Appendix S1: `r s_figure("raw-biplot")`B, D).

@@

Our study integrates diverse, thoroughly collected data sets to characterize the spatial biodiversity of Japanese ferns and gain insight into the processes generating these patterns.
We conclude by summarizing some caveats that must be kept in mind when interpreting our results, while also suggesting avenues for future research.
